Compositional Equivalences Based on Open pNets

### 2022-10-18 2016-Weak-FH-Bisim/WeakBisim.tex
version without symbol change
passe eric+Rabea

\RAB{In the same vein as for context systems \cite{Larsen:91}, pNets is a formalism for modular and possibly incomplete  description of concurrent systems. The two formalisms are however different as the theory of contexts relies on a form of rewrite rules, while  pNets rely on parametric automata to express the system behaviour.  }
\TODO{Rabea: look for papers by Larsen end Liu on context systems, and Larsen and Huttel on static constructs}

@@

We first specify in terms of open transition, what it means for an action to be non-observable. We first define (in Definition~\ref{def:Non-ObsTau}) systems that cannot observe $\tau$ actions of sub-systems; namely pNets that cannot change their state, or emit an observable action when one of its holes emits a $\tau$ action.
\TODO{Eric: discuss rel with Vaandrager - ou pas !!}

@@

Thus, we let $\gamma$ range over sequences of action terms and use $\dotcup$ as the concatenation operator that appends sequences of action terms: given two sequences of action terms  $\gamma\oplus\gamma'$ concatenates the two sequences. The operation is lifted to indexed sets of sequences:   at each index $i$, $\set {\gamma_1}\oplus \set {\gamma_2}$ concatenates the sequences of actions at index $i$ of $\set{\gamma_1}$ and the one at index $i$ of $\set {\gamma_2}$\footnote{One of the two sequences is empty when $i\not \in \dom(\set{\gamma_1})$ or $i\not \in \dom(\set{\gamma_2})$ .}. $[a]$ denotes a sequence with a single element.

@@

We are now able to build a weak open automaton from an open automaton. This is done in a way that resembles the process of $\tau$ saturation: we add  $\tau$ open transitions before or after another open transition, regardless of whether it is observable or not.
\begin{definition}[Building a weak open automaton]\label{def:buildweakOT}~\\
  Let $A = \mylangle J,\mathcal{S},s_0,V,\mathcal{T}\myrangle$ be an open automaton. 
The weak open automaton \emph{derived} from $A$ is an open automaton  $\mylangle J,\mathcal{S},s_0,V,\WT\myrangle$ where $\WT$ is derived from $\mathcal{T}$ by saturation, applying the rules of Figure~\ref{fig:WT}:
%\noindent{\bf Invisible $\tau$ transitions:}
\begin{figure}[tb]
\begin{mathpar}
 {\openrule
         {
           \emptyset,\symb{True},\Id(V)}
         {s \OTWeakarrow {\tau} s} \in \WT  \quad \WTUn}
\qquad\quad
\text{and}
%\noindent{\bf Other transitions:}
\qquad\quad
\inferrule{
 \openrule
         {
           \set{\beta},\Pred,\Post}
         {s \OTarrow {\alpha} s'} \in \mathcal{T}
} 
{ \openrule
         {
           \vis{\set{\beta}}\!,\Pred,\Post
				 } {s \OTWeakarrow {\alpha} s'} \in \WT
} \WTDeux
%\inferrule{
% \openrule
%         {
%           \set{\beta},\Pred,\Post}
%         {s \OTarrow {\tau} s'} \in \mathcal{T}
%\\
% \openrule
%         {
%           \set{\gamma},\Pred\,',({x_k\gets e_k})^{k\in K}   }
%         {s' \OTWeakarrow {\tau} s''} \in \WT
%}
%{ \openrule
%         {
%           \vis{\set{\beta}}\dotcup\set{\gamma},\Pred\land\Pred\,'\subst{\Post\,},
%				({x_k\gets (e_k\subst{\Post\,})})^{k\in K} } 
%         {s \OTWeakarrow {\tau} s''} \in \WT
%}
 \end{mathpar}
 and
%\noindent{\bf Other transitions:}
\begin{mathpar}
\inferrule {\openrule
         {
           \set{\gamma_1},\Pred_1,\Post_1   }
         {s \OTWeakarrow {\tau} s_1} \in \WT
\qquad
\openrule
         {
           \set{\gamma_2},\Pred_2,\Post_2  }
         {s_1 \OTWeakarrow {\alpha} s_2} \in \WT
\\\\
\openrule
         {
           \set{\gamma_3},\Pred_3,\Post_3  }
         {s_2 \OTWeakarrow {\tau} s'} \in\WT
\\
\set{\gamma}=\set{\gamma_1}\dotcup \set{\gamma_2}\subst{\Post_1}\oplus\set{\gamma_3}\subst{\Post_2\shortodot \Post_1}\\
\alpha'=\alpha\subst{\Post_1}\\
\Pred=\Pred_1\land\Pred_2\subst{\Post_1}\land \Pred_3\subst{\Post_2\shortodot\Post_1}
}
{
\openrule
         {\set{\gamma}
           ,
		\Pred,
				\Post_3\shortodot\Post_2\shortodot\Post_1} 
         {s \OTWeakarrow {\alpha'} s'} \in\WT
} \WTTrois
\end{mathpar}
\caption{Weak transition definition}\label{fig:WT}
\end{figure}
\end{definition}
Rule $\WTUn$ states that it is always possible to perform a non-observable transition, where the state is unchanged and the holes perform no action. Rule~$\WTDeux$ states that each open transition is a weak open transition. Finally,  Rule~$\WTTrois$ allows any number of $\tau$ transitions before or after any weak open transition. This rule carefully composes predicates, effects, and actions of the holes. Indeed, predicate $\Pred_2$ manipulates variables of $s_1$ that result from the first weak open transition. Their values thus depend on the initial state but also on the effect (as a substitution function $\Post_1$) of the first weak open transition. In the same manner, $\Pred_3$ must be applied the substitution defined by the composition $\Post_2\shortodot\Post_1$. Similarly, effects on variables must be applied to obtain the global effect of the composed weak open transition, to observable actions of the holes, and to the global action of the weak open transition.

@@

\begin{proof} 
 Suppose that we have: 
\begin{mathpar}
 P[Q]_{j_0} \models \openrule
			{
				{\gamma_j^{j\in J}}, \Pred,  
				\Post}
			{\ostate{s_l^{l\in L}} \OTWeakarrow {\alpha}
				\ostate{s_l'^{\, l\in L}}} 
\end{mathpar}
By Lemma \ref{lem-rel-OT-WOT} this implies the following: \\			
$ \forall p\! \in [0..m_1]\,		
	P[Q]_{j_0}\!\models		
\openrule
    {
       \set{\beta_{1p}},\Pred_{1p},\Post_{1p}   }
         {\ostate{s^{l\in L}_{pl}} \OTarrow {\tau} \ostate{s^{l\in L}_{(p+1)l}}}$,  $\qquad  P[Q]_{j_0}	\models \openrule
         {
           \set{\beta_2},\Pred_2,\Post_2 }
         {\ostate{t_l^{l\in L}} \OTarrow {\alpha'} \ostate{{t'}_l^{l\in L}}}
$ 
\\
and 
 $\forall p \in [0..m_2] \,\, P[Q]_{j_0}\!\models \openrule
         {
           \set{\beta_{3p}},\Pred_{3p}, \Post_{3p}   }
         {\ostate{u^{l\in L}_{pl}} \OTarrow {\tau} \ostate{u^{l\in L}_{(p+1)l}}}
$\\
where\\
{\small \begin{align*}
\forall l\in L.\,&s_l=s_{0 l} \wedge s_{(m_1+1) l}=t_l \wedge t'_l = u_{0 l} \wedge u_{(m_2+1) l}=s'_l\\
\alpha=&\alpha'{\subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}}\\
\gamma_j^{j\in J}=&\mybigoplus_{i=0}^{m_1} \vis{\set{\beta_{1i}}\subst{\bigodot_{j=i-1}^{0}\!\! \Post_{1j}}}\oplus\\&  \vis{\set{\beta_2}\subst{\bigodot_{j=m_1}^{0}\Post_{1j}}} \oplus  
{\bigoplus_{i=0}^{m_2} \vis{\set{\beta_{3i}} \subst{\bigodot_{j=i-1}^{0}\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}} }}
\\
\Pred=&\bigwedge_{i=0}^{m_1}\Pred_{1i}\subst{\bigodot_{j=i-1}^{0}\!\!\Post_{1j}}\land\Pred_2 \subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}\land\\&
\bigwedge_{i=0}^{m_2}\Pred_{3i}\subst{\bigodot_{j=i-1}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}\\
\Post=&\bigodot_{j=m_2}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}
\end{align*}}

>>>>

\RAB{In the same vein as for context systems \cite{Larsen:91}, pNets is a formalism for modular and possibly incomplete  description of concurrent systems. The two formalisms are however different as the theory of contexts relies on a form of rewrite rules, while  pNets rely on parametric automata to express the system behaviour.  }
%\TODO{Rabea: look for papers by Larsen end Liu on context systems, and Larsen and Huttel on static constructs}

@@

We first specify in terms of open transition, what it means for an action to be non-observable. We first define (in Definition~\ref{def:Non-ObsTau}) systems that cannot observe $\tau$ actions of sub-systems; namely pNets that cannot change their state, or emit an observable action when one of its holes emits a $\tau$ action.
%\TODO{Eric: discuss rel with Vaandrager - ou pas !!}

@@

Thus, we let $\gamma$ range over sequences of action terms and use $\oplus$ as the concatenation operator that appends sequences of action terms: given two sequences of action terms  $\gamma\oplus\gamma'$ concatenates the two sequences. The operation is lifted to indexed sets of sequences:   at each index $i$, $\set {\gamma_1}\oplus \set {\gamma_2}$ concatenates the sequences of actions at index $i$ of $\set{\gamma_1}$ and the one at index $i$ of $\set {\gamma_2}$\footnote{One of the two sequences is empty when $i\not \in \dom(\set{\gamma_1})$ or $i\not \in \dom(\set{\gamma_2})$ .}. $[a]$ denotes a sequence with a single element.

@@

We are now able to build a weak open automaton from an open automaton. This is done in a way that resembles the process of $\tau$ saturation: we add  $\tau$ open transitions before or after another open transition, regardless of whether it is observable or not.
\begin{definition}[Building a weak open automaton]\label{def:buildweakOT}~\\
  Let $A = \mylangle J,\mathcal{S},s_0,V,\mathcal{T}\myrangle$ be an open automaton. 
The weak open automaton \emph{derived} from $A$ is an open automaton  $\mylangle J,\mathcal{S},s_0,V,\WT\myrangle$ where $\WT$ is derived from $\mathcal{T}$ by saturation, applying the rules of Figure~\ref{fig:WT}:
%\noindent{\bf Invisible $\tau$ transitions:}
\begin{figure}[tb]
\begin{mathpar}
 {\openrule
         {
           \emptyset,\symb{True},\Id(V)}
         {s \OTWeakarrow {\tau} s} \in \WT  \quad \WTUn}
\qquad\quad
\text{and}
%\noindent{\bf Other transitions:}
\qquad\quad
\inferrule{
 \openrule
         {
           \set{\beta},\Pred,\Post}
         {s \OTarrow {\alpha} s'} \in \mathcal{T}
} 
{ \openrule
         {
           \vis{\set{\beta}}\!,\Pred,\Post
				 } {s \OTWeakarrow {\alpha} s'} \in \WT
} \WTDeux
%\inferrule{
% \openrule
%         {
%           \set{\beta},\Pred,\Post}
%         {s \OTarrow {\tau} s'} \in \mathcal{T}
%\\
% \openrule
%         {
%           \set{\gamma},\Pred\,',({x_k\gets e_k})^{k\in K}   }
%         {s' \OTWeakarrow {\tau} s''} \in \WT
%}
%{ \openrule
%         {
%           \vis{\set{\beta}}\dotcup\set{\gamma},\Pred\land\Pred\,'\subst{\Post\,},
%				({x_k\gets (e_k\subst{\Post\,})})^{k\in K} } 
%         {s \OTWeakarrow {\tau} s''} \in \WT
%}
 \end{mathpar}
 and
%\noindent{\bf Other transitions:}
\begin{mathpar}
\inferrule {\openrule
         {
           \set{\gamma_1},\Pred_1,\Post_1   }
         {s \OTWeakarrow {\tau} s_1} \in \WT
\qquad
\openrule
         {
           \set{\gamma_2},\Pred_2,\Post_2  }
         {s_1 \OTWeakarrow {\alpha} s_2} \in \WT
\\\\
\openrule
         {
           \set{\gamma_3},\Pred_3,\Post_3  }
         {s_2 \OTWeakarrow {\tau} s'} \in\WT
\\
\set{\gamma}=\set{\gamma_1}\oplus \set{\gamma_2}\subst{\Post_1}\oplus\set{\gamma_3}\subst{\Post_2\shortodot \Post_1}\\
\alpha'=\alpha\subst{\Post_1}\\
\Pred=\Pred_1\land\Pred_2\subst{\Post_1}\land \Pred_3\subst{\Post_2\shortodot\Post_1}
}
{
\openrule
         {\set{\gamma}
           ,
		\Pred,
				\Post_3\shortodot\Post_2\shortodot\Post_1} 
         {s \OTWeakarrow {\alpha'} s'} \in\WT
} \WTTrois
\end{mathpar}
\caption{Weak transition definition}\label{fig:WT}
\end{figure}
\end{definition}
Rule $\WTUn$ states that it is always possible to perform a non-observable transition, where the state is unchanged and the holes perform no action. Rule~$\WTDeux$ states that each open transition is a weak open transition. Finally,  Rule~$\WTTrois$ allows any number of $\tau$ transitions before or after any weak open transition. This rule carefully composes predicates, effects, and actions of the holes. Indeed, predicate $\Pred_2$ manipulates variables of $s_1$ that result from the first weak open transition. Their values thus depend on the initial state but also on the effect (as a substitution function $\Post_1$) of the first weak open transition. In the same manner, $\Pred_3$ must be applied the substitution defined by the composition $\Post_2\shortodot\Post_1$. Similarly, effects on variables must be applied to obtain the global effect of the composed weak open transition, to observable actions of the holes, and to the global action of the weak open transition.

@@

\begin{proof} 
 Suppose that we have: 
\begin{mathpar}
 P[Q]_{j_0} \models \openrule
			{
				{\gamma_j^{j\in J}}, \Pred,  
				\Post}
			{\ostate{s_l^{l\in L}} \OTWeakarrow {\alpha}
				\ostate{s_l'^{\, l\in L}}} 
\end{mathpar}
By Lemma \ref{lem-rel-OT-WOT} this implies the following: \\			
$ \forall p\! \in [0..m_1]\,		
	P[Q]_{j_0}\!\models		
\openrule
    {
       \set{\beta_{1p}},\Pred_{1p},\Post_{1p}   }
         {\ostate{s^{l\in L}_{pl}} \OTarrow {\tau} \ostate{s^{l\in L}_{(p+1)l}}}$,  $\qquad  P[Q]_{j_0}	\models \openrule
         {
           \set{\beta_2},\Pred_2,\Post_2 }
         {\ostate{t_l^{l\in L}} \OTarrow {\alpha'} \ostate{{t'}_l^{l\in L}}}
$ 
\\
and 
 $\forall p \in [0..m_2] \,\, P[Q]_{j_0}\!\models \openrule
         {
           \set{\beta_{3p}},\Pred_{3p}, \Post_{3p}   }
         {\ostate{u^{l\in L}_{pl}} \OTarrow {\tau} \ostate{u^{l\in L}_{(p+1)l}}}
$\\
where\\
{\small \begin{align*}
\forall l\in L.\,&s_l=s_{0 l} \wedge s_{(m_1+1) l}=t_l \wedge t'_l = u_{0 l} \wedge u_{(m_2+1) l}=s'_l\\
\alpha=&\alpha'{\subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}}\\
\gamma_j^{j\in J}=&\bigoplus_{i=0}^{m_1} \vis{\set{\beta_{1i}}\subst{\bigodot_{j=i-1}^{0}\!\! \Post_{1j}}}\oplus\\&  \vis{\set{\beta_2}\subst{\bigodot_{j=m_1}^{0}\Post_{1j}}} \oplus  
{\bigoplus_{i=0}^{m_2} \vis{\set{\beta_{3i}} \subst{\bigodot_{j=i-1}^{0}\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}} }}
\\
\Pred=&\bigwedge_{i=0}^{m_1}\Pred_{1i}\subst{\bigodot_{j=i-1}^{0}\!\!\Post_{1j}}\land\Pred_2 \subst{\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}\land\\&
\bigwedge_{i=0}^{m_2}\Pred_{3i}\subst{\bigodot_{j=i-1}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}}\\
\Post=&\bigodot_{j=m_2}^{0}\!\!\Post_{3j}\shortodot\Post_2\shortodot\bigodot_{j=m_1}^{0}\!\!\Post_{1j}
\end{align*}}
